A Bayesian approach to uncertainty in word embedding bias estimation

### 2023-09-06 paper/CLsubmission/bayesBiasCL.tex
Adds review changes

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings, such as
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}:
\emph{Word Embedding Association Test} (\textsf{WEAT}) \citep{Caliskan2017semanticsBiases}, and
\emph{Mean Average Cosine Distance} (\textsf{MAC}) \citep{Manzini2019blackToCriminal}. We first
explain how these measures are supposed to work. Then we argue that they
are problematic for various reasons---the key one being that by
pre-averaging data they manufacture false confidence, which we
illustrate in terms of simulations showing that the measures often
suggest the existence of bias even if by design it is non-existent in a
simulated dataset.

@@

\begin{align}
\label{eq:stAB}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align} \normalsize \noindent then, the association difference
between \(A\) a \(B\) is:

@@

Another issue to consider is the selection of attributes for bias
measurement. The word lists used in the literature are often fairly
small (5-50). The papers in the field do employ statistical tests to
measure the uncertainty involved and do make claims of statistical
significance. Yet, we will later on argue that these method are not
proper for the goal at hand. By using Bayesian methods we will show that
a more appropriate use of statistical methods leads to estimates of
uncertainty which suggest that larger word lists would be advisable.

@@

Bayesian analyses and visualizations thereof can be also handy when it
comes to the investigation of the effect that debiasing has on the
embedding space. In Figures \ref{fig:empiricalPriorToDebiasing} and
\ref{fig:empiricalDebiased} we see an example of two visualizations
depicting the difference in means with 89\% highest posterior density
intervals before and after applying debiasing (the remaining
visualizations are in the Appendix).

@@

\item \citet{Guo2021CEAT} introduce the Contextualized Embedding Association Test
  Intersectional, meant to apply to dynamic word embeddings and,
  importantly develop methods for intersectional bias detection. The
  measure is a generalization of the WEAT method. The authors do inspect
  a distribution of effect sizes that arises from the consideration of
  various possible contexts, but they continue to standardize the
  difference in averaged means and use a single-number summary: the
  weighted mean of the effect sizes thus understood. The method,
  admittedly, deserves further evaluation which goes beyond the scope of
  this paper.
\end{itemize}

@@

\label{appendix:manzini_word_lists} The lists from {[}13{]} are
available here:

>>>>

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings, such as
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\citet{Pennington2014Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\citet{Rabinovich2018Reddit}\footnote{Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}:
\emph{Word Embedding Association Test} (\textsf{WEAT}) \citep{Caliskan2017semanticsBiases}, and
\emph{Mean Average Cosine Distance} (\textsf{MAC}) \citep{Manzini2019blackToCriminal}. We first
explain how these measures are supposed to work. Then we argue that they
are problematic for various reasons---the key one being that by
pre-averaging data they manufacture false confidence, which we
illustrate in terms of simulations showing that the measures often
suggest the existence of bias even if by design it is non-existent in a
simulated dataset.

@@

\begin{align}
\label{eq:stAB}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align} \normalsize \noindent then, the association difference
between \(A\) and \(B\) is:

@@

Another issue to consider is the selection of attributes for bias
measurement. The word lists used in the literature are often fairly
small (5-50). The papers in the field do employ statistical tests to
measure the uncertainty involved and do make claims of statistical
significance. Yet, we will later on argue that these methods are not
proper for the goal at hand. By using Bayesian methods we will show that
a more appropriate use of statistical methods leads to estimates of
uncertainty which suggest that larger word lists would be advisable.

@@

Bayesian analyses and visualizations thereof can be also handy when it
comes to the investigation of the effect that debiasing has on the
embedding space. We used the embeddings that were debiased using \textit{hard} mode debiasing from \citet{Manzini2019blackToCriminal}. In Figures \ref{fig:empiricalPriorToDebiasing} and
\ref{fig:empiricalDebiased} we see an example of two visualizations
depicting the difference in means with 89\% highest posterior density
intervals before and after applying debiasing (the remaining
visualizations are in the Appendix).

@@

\item \citet{Guo2021CEAT} introduce the Contextualized Embedding Association Test
  Intersectional, meant to apply to dynamic word embeddings and,
  importantly develop methods for intersectional bias detection. The
  measure is a generalization of the WEAT method. The authors do inspect
  a distribution of effect sizes that arises from the consideration of
  various possible contexts, but they continue to standardize the
  difference in averaged means and use a single-number summary: the
  weighted mean of the effect sizes thus understood. The method,
  admittedly, deserves further evaluation which goes beyond the scope of
  this paper.

\item \citet{Lum2022Debiasing} employ ..

\item \citet{Ethayarajh2019understanding} points out known limitations to the use of WEAT, namely its sensitivity to the word frequency of the given data and the effect of the choice of word lists. As a result, WEAT systematically overestimates the bias present in the data. The authors propose Relational Inner Product Association (RIPA) to measure association in word embeddings. This method uses the scalar projection of a word vector onto a one-dimensional bias subspace (which is found by employing approach from \citet{Bolukbasi2016man}). RIPA method can be used for those embedding models that do implicitly matrix factorization and contain a co-occurrence statistic. Unlike RIPA, our Bayesian approach can be used for contextual models.

\item \citet{zhang2020robustness} shows the limitations of methods that use gender word pairs to detect bias in the embeddings. They claim that using analogies to detect bias may not necessarily reflect societal bias, but rather the similarity between the words. They conduct experiments where they evaluate four popular bias mesures: Direct Bias (DB), Word Association (WA), Neighbourhood Bias Metric (NBM), and Relational Inner Product Association (RIPA). They show that these measures are not robust to changing either the base pair or the form of a word used. It is also difficult to compare measures because, although they are often presented as better alternatives, they all have similar limitations.

\item \citet{Du2021Assessing} investigates the reliability of bias measures. They find that the bias measure can depend on the way the embedding was obtained. The words that are not frequent in the training data may not have a reliable vector representation that can be used for bias estimation.

\item \citet{Goldfarb2021BiasNotCorrelate} conducts research to understand the relationship between intrinsic bias, which concerns the representation of embeddings, and extrinsic bias, which concerns downstream tasks. The authors seek to answer the question of whether intrinsic bias correlates with extrinsic bias. They examine the intrinsic bias metric WEAT against the extrinsic bias metrics Equality of Opportunity and Predictive Parity and conclude that the correlation is limited. This may further confirm that WEAT is not the most appropriate approach to account for bias in a model.

\item \citet{Spliethover2021BiasSilhouette} propose Bias Silhouette Analysis (BSA), a method for assessing the quality of metrics that measure bias in word embedding models based on word lists. BSA quantifies how robust the word lists used are and how accurate the metrics are with respect to an unbiased and a biased reference model. They examine the Embedding Coherence Test (ECT), the Relative Negative Sentiment Bias (RNSB), and the Word Embedding Association Test (WEAT). Their conclusion is that none of these metrics can reliably discriminate between biased and non-biased models in all cases.

\end{itemize}

@@

\label{appendix:manzini_word_lists} The lists from \citet{Manzini2019blackToCriminal} are
available here:
